### 2021-07-05 rad-icfp-2021/generalization.tex
fixing shift1
finishing some examples

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

@@

\begin{center}
\begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
        && !let A0 = scanl (x,y.e1) y1 A in! \\
        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
        && !let A3 = scanr * 1 A1! \\
        && !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
        && !Y((x1,$\ldots$,xn),0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

>>>>

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

@@

\begin{center}
\begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
        && !let A0 = shift1R (scanl (x,y.e1) y1 A) in! \\
        && !let A1 = shiftL (map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A) in! \\
        && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A in! \\
        && !let A3 = scanr * 1 A1 in! \\
        && !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
        && !Y((x1,$\ldots$,xn),0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

### 2021-07-05 rad-icfp-2021/intro.tex
minor syntax
small fixes
finishing to erase shift1
finishing some examples

This raises several issues that we deal with throughout the paper. First, there is a tradeoff to reach between a general expressive language and a domain specific one. The latter usually has more static information and a specific representation that lends itself to better optimizations.
Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
This makes these systems harder to prove correct. The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency. In addition, we would like something easily provably efficient. 
A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
    \item a detailed comparison with other work and some limitations of our work (\S\ref{sec:conclusion}) \YK{Is it really necessary to specify this point in the contributions? (as it is supposed to be part of every paper)}
\end{itemize}

@@

\begin{example}[First-order term]
On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$, ..., y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$, ..., y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
        !w$_2'$(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !shift1 [v$_1$,...,v$_n$] = [1,v$_1$,...,v$_{n-1}$]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

>>>>

This raises several issues that we deal with throughout the paper. First, there is a tradeoff to reach between a general expressive language and a domain specific one. The latter usually has more static information and a specific representation that lends itself to better optimizations.
Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
This makes these systems harder to prove correct. The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency. In addition, we would like something easily provably efficient. 
A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
\end{itemize}

@@

\begin{example}[First-order term]
On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$,$\ldots$, y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$,$\ldots$, y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
        !w$_2'$(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))! \\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !scanl! and !scanr! are respectively scan left and scan right. 
These return all the intermediate result of a fold in an array.
!shift1L [v$_1$,$\ldots$,v$_n$] \defeq [v$_2$,$\ldots$,v$_{n}$]! and 
!shift1R [v$_1$,$\ldots$,v$_n$] \defeq [v$_1$,$\ldots$,v$_{n-1}$]! are shift operators.
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

### 2021-07-05 rad-icfp-2021/purerad.tex
finishing some examples

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$!) \\
        =& !let w$_1$,Y$_1$=!\\
        & \quad\quad !let y$_1$1,Y$_1$1= <x$_1$, fun (y$_1$,y$_2$,y$_3$,z) -> Y(y$_1$+z,y$_2$,y$_3$)> in! \\
        & \quad\quad !let y$_{12}$,Y$_{12}$= <x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{11}$(y$_{1}$,y$_{2}$+z,y$_{3}$,y$_{4}$)> in! \\
        & \quad\quad !<y$_{11}$ * y$_{12}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{12}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{12}$*z,y$_{11}$*z) > in! \\
        & !let w$_{2}$,Y$_{2}$=!\\
        & \quad\quad !let y$_{21}$,Y$_{21}$= <w$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & \quad\quad !let y$_{22}$,Y$_{22}$= <x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$,z) -> Y$_{11}$(y$_{1}$+z,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$)> in! \\
        & \quad\quad !<y$_{21}$ * y$_{22}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{22}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{22}$*z,y$_{21}$*z) > in! \\
        & !let y,Y$_{3}$= <w, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
    \end{tabular}
    \medskip

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
        =& !let y,Y$_{1}$= <0, fun (X,z) -> Y(X)> in! \\
        & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
        & !let A$_{0}$= shift1R (scanl * y B) in!\\
        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ B) in!\\
        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ B in!\\
        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
        & !<prod(B), fun (X,z) -> Y$_{2}$(X,0,map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
    \end{tabular}
\end{example}

@@

TODO: redo
Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
which we initialize at $1$ to run the backpropagation. 
All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.

@@

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
$\transto$ 
!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
$\transto$
!fun (x$_{1}$,$\ldots$,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],$\ldots$,f$_n$[e$_n$/y$_n$])!

@@

\begin{example}
    The gradient of the terms from the introduction reduces to

    \begin{tabular}{c l}
        & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
        =& TODO \\
    \end{tabular}
    \medskip

    \begin{tabular}{c l}
        & $\grad_\Gamma$(!prod(A)!) \\
        =& TODO \\
    \end{tabular}
\end{example}

>>>>

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$!) \\
        =& !let w$_1$,Y$_1$=!\\
        & \quad\quad !let y$_{11}$,Y$_{11}$= <x$_1$, fun (y$_1$,y$_2$,y$_3$,z) -> Y(y$_1$+z,y$_2$,y$_3$)> in! \\
        & \quad\quad !let y$_{12}$,Y$_{12}$= <x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{11}$(y$_{1}$,y$_{2}$+z,y$_{3}$,y$_{4}$)> in! \\
        & \quad\quad !<y$_{11}$ * y$_{12}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{12}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{12}$*z,y$_{11}$*z) > in! \\
        & !let w$_{2}$,Y$_{2}$=!\\
        & \quad\quad !let y$_{21}$,Y$_{21}$= <w$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_1$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & \quad\quad !let y$_{22}$,Y$_{22}$= <x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$,z) -> Y$_{21}$(y$_{1}$+z,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$)> in! \\
        & \quad\quad !<y$_{21}$ * y$_{22}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{22}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{22}$*z,y$_{21}$*z) > in! \\
        & !let y,Y$_{3}$= <w, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
    \end{tabular}
    \medskip

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
        =& !let y,Y$_{1}$= <1, fun (X,z) -> Y(X)> in! \\
        & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y$_{1}$(X+Z,x)> in!\\
        & !let A$_{0}$= shift1R (scanl * y B) in!\\
        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ B) in!\\
        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ B in!\\
        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
        & !<prod(B), fun (X,z) -> Y$_{2}$(X,0,map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
    \end{tabular}
\end{example}

@@

Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
which we initialize at $1$ to run the backpropagation. 
All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.

@@

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite

\begin{tabular}{l}
!fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
$\transto$ \\
!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
$\transto$ \\
!fun (x$_{1}$,$\ldots$,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],$\ldots$,f$_n$[e$_n$/y$_n$])!
\end{tabular}

@@

\begin{example}
    After forward substitution and inlining the inner !Y!$_{i}$, the gradient of the terms from the introduction reduces to

    \begin{tabular}{c l}
        & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
        =& !let w$_1$,Y$_1$= <x$_1$ * x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y(y$_{1}$+x$_2$*z,y$_{2}$+x$_1$*z,y$_{3}$) > in! \\
        & !let w$_{2}$,Y$_{2}$= <w$_{1}$ * x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{1}$(y$_{1}$+w$_1$*z,y$_{2}$,y$_{3}$,y$_{4}$+x$_{1}$*z) > in! \\
        & !let y,Y$_{3}$= <w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
    \end{tabular}

After another simplification step, we obtain

\begin{tabular}{c l}
    & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
    =& !let w$_1$= x$_1$ * x$_{2}$ in! \\
        & !let w$_{2}$= w$_{1}$ * x$_{1}$ in! \\
        & !<w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> !\\
        & \quad\quad\quad!Y(y$_{1}$+w$_1$*(y$_4$+z)+x$_2$*(y$_4$+x$_1$*(y$_4$+z)),y$_{2}$+x$_1$*(y$_4$+x$_1$*(y$_4$+z)),y$_{3}$) >!
\end{tabular}

Similarly, for the gradient of !prod(A)! we obtain

\begin{tabular}{c l}
        & $\grad_\Gamma$(!prod(A)!) \\
        & !let A$_{0}$= shift1R (scanl * 1 A) in!\\
        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ A) in!\\
        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ A in!\\
        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
        & !<prod(A), fun (X,z) -> Y(X+map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
    \end{tabular}
\end{example}
